---
title: "Oregon Precipitation EDA"
author: "Miles Moran"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

# set WD to the location of this file
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Prerequisite Packages

```{r}
library(ggplot2)
library(dplyr)
library(magrittr)
library(lubridate)

library(ncdf4)      # for loading .nc files
library(maps)
```


## Reading and Viewing the Data

The data is a pared-down version of the Global Historical Climatology Network daily (GHCNd) database maintained by NOAA (with scope limited to just locations in Oregon and just measurements between 1979 and 2017). It contains daily precipitation measurements collected at 208 different weather stations. These are compiled in `rain-daily.rds` can be explored fairly easily:

```{r}

# (1) import contents of rds file
# (2) convert precipitation to mm
# (3) remove observations that are extremely large (>300mm)
# (4) flag observations that are suspiciously large (>200mm and not removed)

rain <- 
    readRDS("../Data/rain-daily.rds") %>% 
    rename(Precip = Precipitation) %>% 
    mutate(
        Precip = Precip/10,
        Precip = ifelse(Precip > 300, NA, Precip),
        Errant = ifelse(Precip > 200 & !is.na(Precip), 1, 0),
        YMD = ymd(paste(Years, Months, Days, sep="-"))
    ) 

rain.monthly <-
    rain %>% 
    group_by(Id, Years, Months) %>% 
    summarize(
        CountRecords  = sum(!is.na(Precip)),
        CountRainyDay = sum(!is.na(Precip) & (Precip > 0)),
        Mean        = mean(Precip, na.rm=TRUE),
        Median      = median(Precip, na.rm=TRUE),
        Std_Dev     = sd(Precip, na.rm=TRUE),
        Quantile_95 = quantile(Precip, 0.95, na.rm=TRUE),
        Quantile_99 = quantile(Precip, 0.95, na.rm=TRUE),
        Max         = max(Precip, na.rm=TRUE) 
    ) %>% 
    ungroup()
                    
rain.USC036 <- filter(rain, Id=="USC00350036")
```

\newpage

## Some Introductory Maps

```{r}
################################################################################
### Empty Map

borders <- 
    map_data("state") %>% 
    filter(region == "oregon") %>% 
    mutate(long=-long)
    
map.oregon <- 
    ggplot(mapping=aes(x=-long, y=lat)) + 
    geom_polygon(aes(group=group), data=borders, fill="white", color="black") + 
    coord_fixed(1.3) + 
    theme_void()

################################################################################
### Map of Station Locations

station.coordinates <- 
    rain %>% 
    group_by(Id) %>% 
    slice(1) %>% 
    ungroup() %>% 
    select(Id, lat, long) 

station.coordinates.highlighted <-
    station.coordinates %>% 
    mutate(isUSC.036 = (Id=="USC00350036"))

map.stations <- 
    map.oregon + 
    geom_point(
        aes(color=!isUSC.036),
        data=station.coordinates.highlighted,
        show.legend=FALSE
    ) + 
    labs(x="", y="", color="Precip. (mm)")

map.stations

ggsave("../Plots/Map-AllStationLocations.png", width=1000, height=500, dpi=100,
       units="px", bg = 'white')
```

\newpage

## Some Introductory Time-Series Plots

```{r}
################################################################################

rain.USC036.new <-
    rain.USC036 %>% 
    filter(YMD > ymd("2010-1-1")) %>% 
    filter(YMD < ymd("2017-1-1")) %>% 
    group_by(Years) %>% 
    mutate(
        YearMax = max(Precip, na.rm=TRUE),
        isYearMax = (Precip == YearMax)
    ) %>% 
    ungroup()

rain.USC036.yrmax <- filter(rain.USC036.new, isYearMax)

years <- filter(rain.USC036.new, Months==1 & Days==1) %>% use_series(YMD)

################################################################################

tsplot.USC036.empty <- 
    ggplot(rain.USC036.new, aes(x=YMD, y=Precip)) + 
    labs(x="Date", y="Precipitation") + 
    scale_y_continuous(limits=c(0,30)) + 
    scale_x_date(limits=c(dmy("01-01-2010"), dmy("01-01-2017")))

# BASIC TSPLOT
tsplot.USC036.empty + 
        geom_line() + geom_point()

# BASIC TSPLOT + YEAR MARKERS
tsplot.USC036.empty + 
        geom_line() + geom_point() + 
        geom_vline(xintercept=years, color="blue", linetype=4)

# BASIC TSPLOT + YEAR MARKERS + ANNUAL MAXIMUM SERIES
tsplot.USC036.empty + 
        geom_line() + geom_point() + 
        geom_vline(xintercept=years, color="blue", linetype=4) + 
        geom_line(color="red", data=rain.USC036.yrmax, show.legend=FALSE) + 
        geom_point(color="red", data=rain.USC036.yrmax, show.legend=FALSE)

ggsave("../Plots/USC00350036-AMS.png", width=1000, height=300, dpi=100, units="px")

# YEAR MARKERS + ANNUAL MAXIMUM SERIES
tsplot.USC036.empty + 
        geom_vline(xintercept=years, color="blue", linetype=4) + 
        geom_line(color="red", data=rain.USC036.yrmax, show.legend=FALSE) + 
        geom_point(color="red", data=rain.USC036.yrmax, show.legend=FALSE) 

# ANNUAL MAXIMUM SERIES ALONE
rain.USC036.new %>% 
    filter(isYearMax) %>% 
    ggplot(aes(x=Years, Precip)) +
        geom_line(color="red") + geom_point(color="red") + 
        labs(x="Date", y="Precipitation") + 
        scale_y_continuous(limits=c(0,30))


################################################################################
```

\newpage

## Combining Map + TS Plot

```{r}

# map.stations
# 
# tsplot.USC036 <- 
#     tsplot.USC036.empty + geom_line() + geom_point()
# 
# p <- gridExtra::grid.arrange(map.stations, tsplot.USC036, ncol=1, nrow=2)
# 
# ggsave("Plots/WinterMaxPcp-Estimates-Stationary.png", plot=p,
#        width=2000, height=1545, units="px")
```

\newpage

## "Histogram" of Daily Precipitation Values

```{r}

rain.USC036 %>% 
    mutate(Precip = case_when(
        (Precip == 0) ~     "0",
        (Precip <  5) ~   "0-5",
        (Precip < 10) ~ "05-10",
        (Precip < 15) ~ "10-15",
        (Precip < 20) ~ "15-20",
        (Precip < 25) ~ "20-25",
        (Precip < 30) ~ "25-30",
        (Precip < 35) ~ "30-35",
        (Precip < 40) ~ "35-40",
        (Precip < 45) ~ "40-45",
        (is.na(Precip)) ~ "NA"
    )) %>% 
    ggplot(aes(x=Precip)) + 
        geom_bar(aes(y=..count..)) + 
        geom_text(aes(label = ..count..), stat = "count", 
                  vjust = -1.5, colour = "black") + 
        geom_point(data=data.frame(
                            Precip=c("20-25","25-30","30-35","35-40","40-45"), 
                            count=c(0,0,0,0,0)),
                   mapping=aes(y=count), cex=0.75) + 
        expand_limits(y=10000) + 
        labs(x="Precipitation (mm)", y="Count (Days)")

ggsave("../Plots/USC00350036-PrecipHist.png", width=1000, height=300, dpi=100, units="px")

```








